Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm: A Case Study

### 2022-05-03 Paper_Draft2/Draft2.Rmd
knit

Due to the proprietary nature of the COMPAS algorithm, it is unknown how exactly these recidivism risk scores are calculated. However, a [sample COMPAS Risk Assessment Survey](https://www.documentcloud.org/documents/2702103-Sample-Risk-Assessment-COMPAS-CORE#document/p5/a296598) has been made publicly available, revealing the algorithm’s input information. @angwin2016machine critiques this survey for using proxy variables for race that do not explicitly factor in a defendant’s race but heavily imply it, allowing Northpointe to claim that their algorithm is free of racial bias. For example, the COMPAS risk assessment survey asks screeners to speculate if a defendant might be affiliated with a gang. It also asks if a defendant has any friends or family members who have been crime victims [@Angwin2016Sample]. Although these questions do not directly ask about race, they do not take into account the pervasive nature of systemic racism that infiltrates every aspect of the lives of marginalized people, thereby indirectly asking about race.

@@

## Equal Opportunity Difference
This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups. The true positive rate is the ratio of true positives to the total number of actual positives for a given group [@GoogleDev].

@@

In-processing mitigates bias in classifiers while building a model. A classifier “is an algorithm that automatically orders or categorizes data into one or more sets” [@baxter2021what]. The in-processing technique we use is the prejudice remover algorithm, which accounts for the fairness metric as part of the input and returns a classifier optimized by that particular metric. In order to do this, we first needed to convert our data frame into a data type called a BinaryLabelDataset. Similarly to pre-processing, we compared the results of our in-processing methods with both the baseline and the pre-processing to gauge which method so far has better individual or group fairness.

>>>>

Due to the proprietary nature of the COMPAS algorithm, it is unknown how exactly these recidivism risk scores are calculated. However, a [sample COMPAS Risk Assessment Survey](https://www.documentcloud.org/documents/2702103-Sample-Risk-Assessment-COMPAS-CORE#document/p5/a296598) has been made publicly available, revealing the algorithm’s input information. @angwin2016machine critiques this survey for using proxy variables for race that do not explicitly factor in a defendant’s race but heavily imply it, allowing Northpointe to claim that their algorithm is free of racial bias. For example, the COMPAS risk assessment survey asks screeners to speculate if a defendant might be affiliated with a gang. It also asks if a defendant has any friends or family members who have been crime victims. Although these questions do not directly ask about race, they do not take into account the pervasive nature of systemic racism that infiltrates every aspect of the lives of marginalized people, thereby indirectly asking about race.

@@

## Equal Opportunity Difference
The [equal opportunity difference](https://developers.google.com/machine-learning/glossary/fairness#e) metric is computed as the difference of true positive rates between the unprivileged and the privileged groups. The true positive rate is the ratio of true positives to the total number of actual positives for a given group.

@@

In-processing mitigates bias in classifiers while building a model. A classifier “is an algorithm that automatically orders or categorizes data into one or more sets” [@baxter2021AI]. The in-processing technique we use is the prejudice remover algorithm, which accounts for the fairness metric as part of the input and returns a classifier optimized by that particular metric. In order to do this, we first needed to convert our data frame into a data type called a BinaryLabelDataset. Similarly to pre-processing, we compared the results of our in-processing methods with both the baseline and the pre-processing to gauge which method so far has better individual or group fairness.

### 2022-05-03 Paper_Draft2/Draft2.tex
knit

Due to the proprietary nature of the COMPAS algorithm, it is unknown how
exactly these recidivism risk scores are calculated. However, a
\href{https://www.documentcloud.org/documents/2702103-Sample-Risk-Assessment-COMPAS-CORE\#document/p5/a296598}{sample
COMPAS Risk Assessment Survey} has been made publicly available,
revealing the algorithm's input information. \citet{angwin2016machine}
critiques this survey for using proxy variables for race that do not
explicitly factor in a defendant's race but heavily imply it, allowing
Northpointe to claim that their algorithm is free of racial bias. For
example, the COMPAS risk assessment survey asks screeners to speculate
if a defendant might be affiliated with a gang. It also asks if a
defendant has any friends or family members who have been crime victims
\citep{Angwin2016Sample}. Although these questions do not directly ask
about race, they do not take into account the pervasive nature of
systemic racism that infiltrates every aspect of the lives of
marginalized people, thereby indirectly asking about race.

@@

This metric is computed as the difference of true positive rates between
the unprivileged and the privileged groups. The true positive rate is
the ratio of true positives to the total number of actual positives for
a given group \citep{GoogleDev}.

@@

In-processing mitigates bias in classifiers while building a model. A
classifier ``is an algorithm that automatically orders or categorizes
data into one or more sets'' \citep{baxter2021what}. The in-processing
technique we use is the prejudice remover algorithm, which accounts for
the fairness metric as part of the input and returns a classifier
optimized by that particular metric. In order to do this, we first
needed to convert our data frame into a data type called a
BinaryLabelDataset. Similarly to pre-processing, we compared the results
of our in-processing methods with both the baseline and the
pre-processing to gauge which method so far has better individual or
group fairness.

>>>>

Due to the proprietary nature of the COMPAS algorithm, it is unknown how
exactly these recidivism risk scores are calculated. However, a
\href{https://www.documentcloud.org/documents/2702103-Sample-Risk-Assessment-COMPAS-CORE\#document/p5/a296598}{sample
COMPAS Risk Assessment Survey} has been made publicly available,
revealing the algorithm's input information. \citet{angwin2016machine}
critiques this survey for using proxy variables for race that do not
explicitly factor in a defendant's race but heavily imply it, allowing
Northpointe to claim that their algorithm is free of racial bias. For
example, the COMPAS risk assessment survey asks screeners to speculate
if a defendant might be affiliated with a gang. It also asks if a
defendant has any friends or family members who have been crime victims.
Although these questions do not directly ask about race, they do not
take into account the pervasive nature of systemic racism that
infiltrates every aspect of the lives of marginalized people, thereby
indirectly asking about race.

@@

The
\href{https://developers.google.com/machine-learning/glossary/fairness\#e}{equal
opportunity difference} metric is computed as the difference of true
positive rates between the unprivileged and the privileged groups. The
true positive rate is the ratio of true positives to the total number of
actual positives for a given group.

@@

In-processing mitigates bias in classifiers while building a model. A
classifier ``is an algorithm that automatically orders or categorizes
data into one or more sets'' \citep{baxter2021AI}. The in-processing
technique we use is the prejudice remover algorithm, which accounts for
the fairness metric as part of the input and returns a classifier
optimized by that particular metric. In order to do this, we first
needed to convert our data frame into a data type called a
BinaryLabelDataset. Similarly to pre-processing, we compared the results
of our in-processing methods with both the baseline and the
pre-processing to gauge which method so far has better individual or
group fairness.
